---
title: "HW 03 - Great Lakes fish datasets"
author: "Insert your name here"
date: "Insert date here"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Academic integrity and honor code

By completing and submitting this homework, I certify that:

- I completed the submitted work myself.
- I did not use generative AI (including ChatGPT, Claude, Gemini, Copilot, Grok, or similar tools) to generate, debug, revise, or interpret code or written answers for this assignment.
- I may use course materials, lecture notes, R documentation/help files, and other resources explicitly approved by the instructors.
- I understand that I am responsible for being able to explain all code and written work that I submit.

**Name:** ______________________________  
**Signature:** ___________________________  
**Date:** _______________________________

# Before you begin

Work through `hw-03-instructions.Rmd`. Keep **all submitted code in this file** and Knit frequently. Ask Dr. Gavin McNicol or TA Kayode Ayegbusi if you are still stuck after checking spelling, parentheses, and Knit errors.

If `WEIGHT` imports as character, parse it before Exercise 5:

```{r parse-weight, eval=FALSE}
fish_stock <- fish_stock %>%
  mutate(WEIGHT = parse_number(WEIGHT))
```

# Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r load-data, message=FALSE}
catch_data <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-06-08/fishing.csv")

fish_stock <- read_csv("data/great-lakes-fish-stock.csv")
```

# Exercises

## Exercise 1 — Select columns

**Goal:** Keep `year`, `lake`, `species`, `region`, and `values` from `catch_data`, and `YEAR`, `MONTH`, `LAKE`, `STATE_PROV`, `SPECIES`, and `WEIGHT` from `fish_stock`.

```{r catch-data-subset}
# Your code here
```

```{r fish-stock-subset}
# Your code here
```

## Exercise 2 — Glimpse the subsets

**Goal:** Inspect both objects after `select()`.

```{r glimpse-subsets}
# glimpse(catch_data) and glimpse(fish_stock)
```

## Exercise 3 — Filter to Chinook Salmon in Lake Michigan

**Goal:** U.S. Total Chinook Salmon catch in Lake Michigan; CHS stocking in MI from MI, WI, IL, and IN.

```{r filter-catch}
catch_data <- catch_data %>% 
  filter(region == "U.S. Total",
         species == "___",
         lake == "___")
```

```{r filter-stock}
fish_stock <- fish_stock %>% 
  filter(___ %in% c("MI", "WI", "IL", "IN"),
         SPECIES == "___",
         LAKE == "MI")
```

## Exercise 4 — Row counts and year ranges

**Goal:** Report `nrow`, `min(year)`, and `max(year)` for each filtered table with inline R.

> Replace this text with three complete sentences that use inline R code.

## Exercise 5 — Annual stocking totals

**Goal:** One row per year in `fish_stock_yearly`, with `STOCK_WEIGHT_YR = sum(WEIGHT, na.rm = TRUE)`.

```{r stock-yearly}
# Your group_by / summarize pipeline here
```

## Harmonize names and units

Set `eval=TRUE` after Exercise 5 runs successfully.

```{r finalize-fish-stock, eval=FALSE}
fish_stock_yearly <- fish_stock_yearly %>% 
  rename(year = YEAR) %>% 
  mutate(stock_weight_yr_lbs = as.integer(STOCK_WEIGHT_YR * 2.205 / 1000)) %>% 
  select(year, stock_weight_yr_lbs)
```

```{r finalize-catch-data, eval=FALSE}
catch_data <- catch_data %>% 
  rename(catch_weight_yr_lbs = values)
```

## Exercise 6 — Six joins

**Goal:** Join on `year` six ways and explain which rows each join keeps.

```{r left-join}
# catch_data %>% left_join(fish_stock_yearly)
```

> Replace this text with 1–2 sentences on `left_join()`.

```{r right-join}
# catch_data %>% right_join(fish_stock_yearly)
```

> Replace this text with 1–2 sentences on `right_join()`.

```{r inner-join}
# catch_data %>% inner_join(fish_stock_yearly)
```

> Replace this text with 1–2 sentences on `inner_join()`.

```{r full-join}
# catch_data %>% full_join(fish_stock_yearly)
```

> Replace this text with 1–2 sentences on `full_join()`.

```{r anti-join-v1}
catch_data %>% 
  anti_join(fish_stock_yearly)
```

> Replace this text with 1–2 sentences on this `anti_join()`.

```{r anti-join-v2}
fish_stock_yearly %>% 
  anti_join(catch_data)
```

> Replace this text with 1–2 sentences on this `anti_join()`.

## Exercise 7 — Pivot to long format

**Goal:** `left_join()` then `pivot_longer()` so one column is `activity` and one is `weight_lbs`.

```{r pivot-longer}
catch_data %>% 
  left_join(___) %>% 
  ___(names_to = "activity",
               values_to = "___",
               cols = 5:6)
```

## Exercise 8 — Line plot of catch and stocking

**Goal:** Extend Exercise 7 with `ggplot()`, `geom_line()`, and `theme_bw()`.

```{r final-visualization, fig.width=7, fig.height=4.5}
catch_data %>% 
  left_join(___) %>% 
  ___(names_to = "activity",
               values_to = "___",
               cols = 5:6) %>% 
  ggplot(aes(x = ___, y = ___, color = ___)) +
  geom_line() +
  theme_bw()
```

**Interpretation (1–3 complete sentences):** What happened to Lake Michigan Chinook Salmon stocking and catch after the early 1990s?

> Replace this text with your answer.

# Submission checklist

- [ ] I signed the academic-integrity statement.
- [ ] My name and date are correct in the YAML.
- [ ] I answered every requested question in complete sentences.
- [ ] Exercise 4 uses inline R code.
- [ ] All six joins have a short narrative.
- [ ] This document Knits without errors.
- [ ] I committed and pushed my final changes to GitHub.
