---
title: "Lab 03 - Great Lakes fish stocking"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Academic integrity and honor code

This lab is designed to help you build fluency with wrangling and visualizing environmental data in R. **Generative AI tools are not permitted for this assignment.** This includes ChatGPT, Claude, Gemini, Copilot, Grok, and similar tools that can generate or revise code, written answers, interpretations, or assignment content.

By completing and submitting this lab, I certify that:

- I completed the submitted work myself.
- I did not use generative AI to generate, debug, revise, or interpret code or written answers for this assignment.
- I may use course materials, lecture notes, R documentation/help files, and other resources explicitly approved by the instructors.
- I understand that I am responsible for being able to explain all code and written work that I submit.

**Name:** ______________________________  
**Signature:** ___________________________  
**Date:** _______________________________

> **Important:** Signing this statement is part of the assignment. The same statement appears in your starter file; complete it there as well.

# Lab overview

You will use the [Great Lakes Fish Stocking database](http://www.glfc.org/fishstocking/index.htm), a centralized record of fish stocked from artificial propagation into the Great Lakes since 1950. The database was developed with fishery professionals from the agencies that stock the lakes. Managers use it to understand the food and habitat needs of stocked predators. It is updated and hosted by the U.S. Fish and Wildlife Service.

```{marginfigure}
A stocking event is one release of a species into a lake by an agency in a given year (and month, when reported).
```

## Learning goals

By the end of this lab, you should be able to:  

  * Describe the size and observational unit of a data frame with inline R.
  * Filter rows with `%in%`, `!is.na()`, and `!=`.
  * Create a new variable with `mutate()`.
  * Compute grouped summaries with `group_by()` and `summarize()`.
  * Combine wrangling and `ggplot2` in one pipeline, including `facet_wrap()`.

## Deliverable

Complete **`lab-03-starter.Rmd`**. Your finished starter file should contain your code, plots, and written responses and should Knit successfully from top to bottom.

**The instructions file tells you what to do; the starter file is where your assignment work belongs, except for completing the honor-code signature above.**

## Reproducibility rule

Throughout EAES 380, your analysis should work **from top to bottom**. Code that works only because you previously ran something in the Console is not reproducible.

**Knit frequently.**

## Before asking for help

First check spelling, parentheses, quotation marks, package loading, chunk order, and whether the document Knits. If you are still stuck, ask Dr. Gavin McNicol or TA Kayode Ayegbusi and show them the code and error message.

# Getting started

If you are reading a knitted copy of these instructions, treat the steps below as review.

- Open the GitHub repository you want to copy and add `/generate` to the URL.
- Under **Owner**, choose `EAES-380`. Do not choose your personal account.
- Name the repository `unit-3-yourlastname` and keep it **Private**.
- Clone it into a New Project from a GitHub Repo in [RStudio Cloud](https://rstudio.cloud).
- Open `lab-03-instructions.Rmd` and Knit it.
- Open `lab-03-starter.Rmd`. **This is the file you will complete.**

# Warm-up: R Markdown + Git

Open `lab-03-starter.Rmd`.

A. Replace `Insert your name here` with your name.
B. Replace `Insert date here` with today's date.
C. Complete and sign the honor-code statement.
D. Knit, then commit with a message such as `Update author and honor code` and push to GitHub.

# Packages and data

```{r load-packages-demo, message = FALSE}
library(tidyverse)
```

```{r load-data-demo, message = FALSE}
fish_stock <- read_csv("data/great-lakes-fish-stock.csv")
```

```{marginfigure}
A few `WEIGHT` cells contain commas or notes such as `/k`. If `WEIGHT` imports as character, run `mutate(WEIGHT = parse_number(WEIGHT))` before the exercises so `mean()` and arithmetic work.
```

Variable descriptions:

- `YEAR`: Year of stocking (YYYY)
- `MONTH`: Month of stocking (not always reported)
- `LAKE`: Great Lake code  
  **ER** Erie; **HU** Huron; **MI** Michigan; **ON** Ontario; **SC** St. Clair; **SU** Superior
- `STATE_PROV`: State or province code  
  **IL** Illinois; **IN** Indiana; **MI** Michigan; **MN** Minnesota; **NY** New York; **OH** Ohio; **ON** Ontario; **PA** Pennsylvania; **WI** Wisconsin
- `SPECIES`: Species code
- `NO_STOCKED`: Number of fish stocked
- `AGEMONTH`: Age in months (not always reported)
- `WEIGHT`: Total weight of fish stocked, in kilograms

# Exercises

For each exercise: read the prompt here, complete the matching section in `lab-03-starter.Rmd`, run the code, write any interpretation in complete sentences, and Knit at the checkpoints.

## Exercise 1 — Size and observational unit

**Goal:** Report the dimensions of `fish_stock` with inline R and name the observational unit.

- How many observations and how many variables are in the dataset? Use inline code rather than typing the digits by hand.
- What does each row represent? Be specific.

## Exercise 2 — Filter to salmon stocking events

**Goal:** Build a new data frame with `%in%`, `!is.na()`, and `!=`.

Create `salmon_stock` that keeps:

- stocking events for `ATS` (Atlantic Salmon), `CHS` (Chinook Salmon), and `COS` (Coho Salmon)
- events that report total weight (`WEIGHT` is not `NA`)
- events in every lake **except** Ontario (`LAKE` is not `"ON"`)

```{r salmon-filter-demo, echo = TRUE}
salmon_stock <- fish_stock %>%
  filter(
    SPECIES %in% c("___", "___", "___"),
    !is.na(___),
    ___ != "ON"
  )
```

Then, in a second labeled chunk, report:

- the number of rows
- the number of variables
- the names and types of the variables in `salmon_stock`

Write 2–3 sentences stating those dimensions and which variables are numeric versus character.

### Exercise 2 checkpoint

🧶 ✅ ⬆️ **Knit, commit, and push.** Suggested commit message: `Complete Exercises 1 and 2`.

## Exercise 3 — Average fish weight in grams

**Goal:** Create a new column with `mutate()`.

Pipe `salmon_stock` into `mutate()` and add a column for the average individual fish weight in **grams**. `WEIGHT` is in kilograms and `NO_STOCKED` is a count, so you need to divide and convert kilograms to grams.

```{r avg-weight-demo, echo = TRUE}
salmon_stock <- ___ %>%
  ___(
    ___ = WEIGHT / NO_STOCKED * ___
  )
```

Use ALL CAPS for the new name if you want to match the other columns. Then write one sentence explaining what `mutate()` is doing here: what it creates, and which columns it uses.

### Exercise 3 checkpoint

🧶 ✅ ⬆️ **Knit, commit, and push.** Suggested commit message: `Complete Exercise 3`.

## Exercise 4 — Lake summaries since 2000

**Goal:** Filter, group, summarize, and arrange.

For years **after 2000**, compute the mean number of salmon stocked per event and the mean total weight per event **by lake**. Arrange the table in descending order of mean number stocked.

```{r salmon-summary-demo, echo = TRUE}
___ %>% 
  group_by(___) %>% 
  filter(YEAR > ___) %>% 
  summarize(
    MEAN_NO_STOCKED = mean(___, na.rm = TRUE),
    MEAN_WEIGHT = mean(___, ___ = ___)
  ) %>% 
  arrange(desc(___))
```

Which lake has the largest mean number of fish per stocking event? Which has the largest mean weight per event?

### Exercise 4 checkpoint

🧶 ✅ ⬆️ **Knit, commit, and push.** Suggested commit message: `Complete Exercise 4`.

## Exercise 5 — Repeat the pipeline for trout

**Goal:** Combine filter, mutate, summarize, and arrange in one pipeline on a new species group.

Trout codes in this file:

- `BKT` Brook Trout
- `BNT` Brown Trout
- `LAT` Lake Trout
- `RBT` Rainbow Trout
- `TRT` Tiger Trout

```{marginfigure}
Brown trout is coded `BNT` in this database, not `BRN`. `BRN` matches zero rows.
```

Start from `fish_stock` (not `salmon_stock`) and, in one unlabeled-object pipeline:

- keep those five trout codes, drop missing `WEIGHT`, and drop Lake Ontario
- create the same average individual weight (grams) column
- summarize mean number stocked and mean total weight by lake (no year filter this time)
- arrange by descending mean number stocked

You do not need to assign the result to a new object. Write one sentence naming the lake with the largest mean number of trout per event.

### Exercise 5 checkpoint

🧶 ✅ ⬆️ **Knit, commit, and push.** Suggested commit message: `Complete Exercise 5`.

## Exercise 6 — Annual stocking of Chinook Salmon and Brook Trout

**Goal:** Summarize by year, species, and lake, then draw a faceted line plot.

Create a faceted line plot of the **annual total number** of Chinook Salmon (`CHS`) and Brook Trout (`BKT`) stocked. Each facet is a lake. Color the two species differently.

Build one pipeline:

- Filter `fish_stock` to `CHS` and `BKT`.
- `group_by()` year, species, **and** lake, then `summarize()` the sum of `NO_STOCKED`.
- Pipe into `ggplot()`. Map year to x, the summed count to y, and species to color.
- Add `geom_line()`, facet by `LAKE`, then `theme_bw()` (or another complete theme) and `labs()`.

Answer in 2–3 complete sentences:

- Which of the two species was stocked at a higher rate overall?
- In which lake was stocking largest?
- In that lake, roughly when did stocking reach its peak?

### Exercise 6 checkpoint

🧶 ✅ ⬆️ Knit, commit, and push with a message such as `Complete Lab 3`. Confirm the latest files are on GitHub.

# Submission checklist

- [ ] I signed the academic-integrity statement in both files.
- [ ] My name and date are correct in the starter YAML.
- [ ] I answered every requested question in complete sentences.
- [ ] Exercise 1 uses inline R code.
- [ ] All required code is in `lab-03-starter.Rmd`, not only the Console.
- [ ] Every chunk I added has a label with no spaces.
- [ ] The document Knits from top to bottom without errors.
- [ ] I committed and pushed the final version to GitHub.
