---
title: "Lab 03 - Great Lakes fish stocking"
author: "Insert your name here"
date: "Insert date here"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Academic integrity and honor code

By completing and submitting this lab, I certify that:

- I completed the submitted work myself.
- I did not use generative AI (including ChatGPT, Claude, Gemini, Copilot, Grok, or similar tools) to generate, debug, revise, or interpret code or written answers for this assignment.
- I may use course materials, lecture notes, R documentation/help files, and other resources explicitly approved by the instructors.
- I understand that I am responsible for being able to explain all code and written work that I submit.

**Name:** ______________________________  
**Signature:** ___________________________  
**Date:** _______________________________

# Before you begin

Work through the exercises using `lab-03-instructions.Rmd` as your guide. Keep **all submitted code in this file**, run the analysis from top to bottom, and Knit frequently.

If `WEIGHT` imports as a character column, parse it before the exercises:

```{r parse-weight, eval=FALSE}
fish_stock <- fish_stock %>%
  mutate(WEIGHT = parse_number(WEIGHT))
```

Set `eval=TRUE` if you need that chunk. Ask Dr. Gavin McNicol or TA Kayode Ayegbusi if you are still stuck after checking spelling, parentheses, and Knit errors.

# Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
```

```{r load-data, message=FALSE}
fish_stock <- read_csv("../data/great-lakes-fish-stock.csv")
```

# Exercises

## Exercise 1 — Size and observational unit

**Goal:** Report how many rows and columns `fish_stock` has, using inline R, and say what one row represents.

> Replace this text with your answer. Use inline R such as `` `r nrow(fish_stock)` `` and `` `r ncol(fish_stock)` ``.

## Exercise 2 — Filter to salmon stocking events

**Goal:** Keep Atlantic, Chinook, and Coho salmon events with a reported weight, excluding Lake Ontario.

```{r filter-salmon, eval = F}
salmon_stock <- fish_stock %>%
  filter(
    SPECIES %in% c("___", "___", "___"),
    !is.na(___),
    ___ != "ON"
  )
```

```{r inspect-salmon}
# Number of rows, number of variables, and variable types for salmon_stock
```

**Interpretation (2–3 complete sentences):** How many rows and variables does `salmon_stock` have, and which variables are numeric versus character?

> Replace this text with your answer.

## Exercise 3 — Average fish weight in grams

**Goal:** Add a column for mean individual weight in grams (`WEIGHT` is in kg).

```{r mutate-avg-weight, eval = F}
salmon_stock <- ___ %>%
  ___(
    ___ = WEIGHT / NO_STOCKED * ___
  )
```

**Interpretation (1 complete sentence):** What is `mutate()` creating here, and which columns does it use?

> Replace this text with your answer.

## Exercise 4 — Lake summaries since 2000

**Goal:** Mean number stocked and mean total weight per event by lake, for `YEAR > 2000`, arranged by descending mean number.

```{r summarize-salmon-lakes, eval = F}
___ %>% 
  group_by(___) %>% 
  filter(YEAR > ___) %>% 
  summarize(
    MEAN_NO_STOCKED = mean(___, na.rm = TRUE),
    MEAN_WEIGHT = mean(___, ___ = ___)
  ) %>% 
  arrange(desc(___))
```

**Interpretation (1–2 complete sentences):** Which lake has the largest mean number of fish per event, and which has the largest mean weight per event?

> Replace this text with your answer.

## Exercise 5 — Trout pipeline

**Goal:** One pipeline from `fish_stock` for `BKT`, `BNT`, `LAT`, `RBT`, and `TRT` (no year filter).

```{r summarize-trout-lakes}
# Your pipeline here
```

**Interpretation (1 complete sentence):** Which lake has the largest mean number of trout per stocking event?

> Replace this text with your answer.

## Exercise 6 — Annual CHS and BKT stocking by lake

**Goal:** Faceted line plot of annual total `NO_STOCKED` for Chinook Salmon and Brook Trout.

```{r plot-chs-bkt, fig.width=8, fig.height=6}
# Your pipeline here
```

**Interpretation (2–3 complete sentences):** Which species was stocked at a higher rate? In which lake was stocking largest? Roughly when did stocking peak in that lake?

> Replace this text with your answer.

# Submission checklist

- [ ] I signed the academic-integrity statement.
- [ ] My name and date are correct in the YAML.
- [ ] I answered every requested question in complete sentences.
- [ ] Exercise 1 uses inline R code.
- [ ] All required code is in this file.
- [ ] This document Knits without errors.
- [ ] I committed and pushed my final changes to GitHub.
